Bumble bee foraging bias

### 2021-01-22 writing/manuscript_v4.Rmd
backup

---
title: "Bumble bee foraging bias"
author:
- Douglas B. Sponsler
- Fabrice Requier
- Katharina Kallnik
- Alice Claßen
- A. Fabienne Maihoff
- Ingolf Steffan-Dewenter
bibliography: refs.bib
output:
  pdf_document: default
  html_document: default
---

@@

@@

@@

@@

Prior to analysis, we aligned each bumble bee visitation sample with corresponding floral survey data. For most samples, floral surveying was conducted jointly with bumble bee visitation observation. In cases where bumble bee visitation was sampled without a joint floral survey, the bumble bee visitation data was paired with the nearest available floral survey data within a maximum distance of 7 days. Bumble bee visitation samples lacking corresponding floral survey data within 7 days were omitted from further analysis. In a few cases where more than one bumble bee visitation date was paired with a given floral survey date, the bumble bee visitation data were pooled.

@@

To quantify cumulative bias per bumble bee species, we used the discrimination (d’) metric of @Bluthgen2006-if, which represents the net deviation of a species’ visitation from the predictions of a null model in which the probability of visitation is purely a function of floral abundance.

#### 2.2.2.1.1 Model 1: dprime ~

### 2.2.2 Modeling visitation probability as a function of floral morphotype and bumble bee tongue length

#### 2.2.2.1 Model 2: visits.bin ~ prop.flower.cover + bb.sp.abund + (1 | k.type.ss) + (1 | site)

#### 2.2.2.2 Model 3: visits.bin ~ prop.flower.cover + bb.sp.abund + (1 | k.type.ss:pbl.w.class2) + (1 | site)

#### 2.2.2.3 Model 4: Ordination with post-hoc fitting of tongue-length

@@

# 3. Results
## 3.1 Model 1
```{r}
knitr::include_graphics("../output/brm.plantsp0_plot1.png")
```

## 3.2 Model 2

## 3.3. Model 3

## 3.4 Model 4

## 3.5 Model 5

## 3.6 Model 6

>>>>

---
title: "Bumble bee foraging bias"
author:
- Douglas B. Sponsler
- Fabrice Requier
- Katharina Kallnik
- Alice Claßen
- A. Fabienne Maihoff
- Ingolf Steffan-Dewenter
bibliography: refs.bib
output:
  pdf_document:
    fig_caption: yes
---

@@

### Study sites

@@

### Bumble bee visitation

@@

visitation sample = site-date of bumble bee visitation observations

While floral visitation by male bumble bees was recorded during visitation sampling, we chose to analyze only visitation by queens and workers.

### Flora surveying

floral survey sample = site-date of floral surveying

@@

Prior to analysis, we aligned each visitation sample with a corresponding floral survey sample. In most cases, floral surveying was conducted jointly with bumble bee observation for each site and date. Occasionally, though, floral surveying could not be completed on the same day as bumble bee observation, resulting in a small temporal offset between visitation samples and floral survey samples. In these cases, visitation samples were paired with the nearest available floral survey data within a maximum distance of 7 days. Visitation samples lacking corresponding floral survey data within 7 days were omitted from further analysis. In a few cases, two consecutive visitation samples at a given site shared the same nearest floral survey sample; in these cases, the visitation samples were first pooled, then aligned to their shared nearest floral survey sample. In rare cases, a given floral species was recorded during visitation sampling but missed during floral surveying. These species were retroactively assigned the minimum floral area value of 0.01 m^2.

@@

To quantify cumulative bias per bumble bee species, we used the discrimination (*d’*) metric of @Bluthgen2006-if, which represents the net deviation of a species’ visitation patterns from the predictions of a null model in which the probability of visitation is purely a function of floral abundance. We calculated *d'* values separately for each visitation sample, with floral abundances supplied by its corresponding floral survey sample.

Preliminary visualization showed that *d'* values tended to become binary (0 or 1) when calculated for bumble bee species for which fewer than 3 individuals were observed in a given visitation sample. This generated a tri-modal distribution of *d'* values, with peaks at 0 and 1 flanking an otherwise approximately normal distribution. To ease model estimation and avoid the influence of the highly uncertain *d'* values calculated based on one or two observations, we chose to filter our data to include for each visitation sample only species that were observed at least 3 times.

The variation of *d'* was analyzed using Bayesian multi-level modeling. We modeled the response variable *d'* as a function of the interaction between elevation and day-of-year, with bumble bee species as a varying intercept term. The continuous predictors elevation and day-of-year were scaled and centered to optimize software performance. <!-- Explain choice of priors (or lack thereof). Maybe also explain the choice of modeling bb.sp as a varying intercept term rather than a fixed effect --> We then specified an alternative model in which the elevation and day of year terms were replaced with a varying intercept term for site. This simpler model was compared to the interaction model via leave-one-out (LOO) cross-validation using the *loo()* function from *brms*, which is a wrapper for a function of the same name from the R package *loo* [@Yao2017aa].

### 2.2.2 Modeling visitation rate as a function of floral morphotype and bumble bee tongue length

The discrimination metric *d'* expresses the overall disproportionality between relative abundance and visitation, but it does not reveal the causes of bias, i.e. the functional relationships between bumble bees and flowers that lead to differential visitation. To explore these functional underpinnings of bumble foraging bias, we analyzed visitation rate as a function of the interaction between floral morphology and bumble bee tongue length, conditioned on the abundance of each flower and bumble bee species.

Bumble bee tongue-length data were obtained from the literature values compiled and summarized by @Arbetman2017-xj. Values for *B. gerstaeckeri* and *B. mendax*, which were not included in @Arbetman2017-xj, were obtained from @Obeso1992-yp and @Durieux2000-uk, respectively. We then binned bumble bee species into three tongue-length classes: short (< 7.5 mm), medium (7.5---10.0 mm), and long (>= 10.0 mm). Short-tongued species included *B. jonellus*, *B. lapidarius*, *B. pratorum*, *B. psithyrus-group*, *B. pyrenaeus*, and *B. terrestris/lucorum*. Medium-tongued species included *B. humilis*, *B. hypnorum*, *B. monticola*, *B. mucidus*, *B. pascuorum*, and *B. wurflenii*. Long-tongued species included *B. gerstaeckeri*, *B. hortorum*, and *B. mendax*.

Floral morphology was summarized using floral morphotypes of @Kugler1970-gb, accessed via the BIOFLOR database [@Klotz2002-uc]. Morphotypes were simplified to the basic categories numbered 0-10: 0 = not applicable (includes wind-pollinated flora such as grasses), 1 = disc- and bowl-shaped flowers, 2 = funnel flowers, 3 = bell-shaped flowers, 4 = stalk-disc flowers, 5 = lip flowers, 6 = flag flowers, 7 = flower heads, 8 = spike flowers, 9 = brush flowers, 10 = trap flowers **(Figure X)**. <!-- I think I would like to include a figure with photographs of flowers exemplifying each morphotype --> In a few cases, a plant species in our data set was not found in the BIOFLOR database. In most cases, morphotypes are constant at the genus level, and missing morphotypes were added based on the morphotypes found for congeners. The genus *Gentiana* consisted of both group 2 and group 4 representatives, so the missing species *Gentiana aspera* and *Gentiana ciliata* were classified based on visual comparison with their classified congeners. In the case of the genus *Phyteuma*, we followed @Neumayer1999-il and classified it as a group 7 flower head.

We modeled visitation using multilevel Bayesian Poisson regression. First, we specified a model focused on comparing visitation rates across morphotypes. Visitation was modeled using varying intercept terms for morphotype, site, and bumble bee species, and fixed terms for the abundances of floral and bumble bee species. Then, we specified a more complex model in which the varying intercept term for morphotype was replaced with one for the interaction of morphotype and bumble bee tongue length class. In both models, the continuous predictors were scaled and centered to optimize software performance. <!-- Should I perhaps have included a fixed interaction of bb abundance and floral abundance? -->

@@

### 2.2.4 Software and code

All analyses were conducted in R v. 3.6 . Discrimination (*d'*) was calculated using the function *dfun()* [@Dormann2011-aa] from the R package *bipartite* [@Dormann2008-ab].Bayesian multilevel models were implemented in *Stan* [@Stan2020aa] via the R package *brms* [@Buerkner2018aa], and visualizations were constructed with the R packages *tidybayes* [@Kay2020aa] and *ggdist* [Kay2020ab]. Ordination analyses were performed with the R package *vegan* [@Oksanen2019aa] and visualized using the R packages *ggplot2* [@Wickham2016aa] and *ggvegan* [@Simpson2019aa]. Data handling was performed the *tidyverse* suite of R packages [Wickham2019aa]. Code and model diagnostics are available in Appendix X.

# 3. Results
## 3.1 Cumulative bias (d')

LOO model comparison favored the simpler model that used a varying intercept effect for site in lieu of the elevation:day-of-year interaction term (elpd_diff = -104). The estimated global mean of *d'* was 0.38 (0.33---0.43 95% CI). Variation of bumble bee species from the global mean of *d'* was modest, with an estimated standard deviation of 0.05 (0.02---0.08 95% CI) and substantial overlap between species in their posterior distributions **(Figure XA)**. Spatial variation in *d'* was somewhat more pronounced, with an estimated standard deviation of 0.08 (0.06---0.11 95% CI) **(Figure XB)**.

See supplemental material for the details of model specification and evaluation.

```{r fig1, echo = FALSE,  fig.cap = paste("Posterior conditional mean discrimination (d') by bumble bee species and site. Half-eye intervals depict 66\\% and 95\\% CIs.")}
knitr::include_graphics("../output/brm_dprime.sp01_plot1.png")
```

## 3.2 Explaining bias via analysis of visitation rate by flower morphotype and bee tongue length

Visitation rates varied markedly across morphotypes **(Figure X)**. Flag, lip, and bell flowers had the highest visitation rates, with broadly overlapping posterior distributions. Funnel flowers and bell flowers had the next highest visitation rates, followed by stalk-disc flowers and then disc flowers. Unclassified flowers, which are mainly wind-pollinated species, had the lowest visitation rates.

```{r fig2, echo = FALSE,  fig.cap = paste("Posterior conditional mean visits (log-scale) by Kugler morphotype. Half-eye plots depict 66\\% and 95\\% CIs.")}
knitr::include_graphics("../output/brm_pvis00_plot1.png")
```

Visitation also varied strongly within morphotypes across bumble bee tongue length classes **(Figure X)**. Disc flowers (group 1), bell flowers (group 3), and unclassified flowers (group 0) were visited chiefly by short-tongued bumble bee species; visitation of these morphotypes by medium- and long-tongued species was much lower, and there was little difference in visitation rates between medium- and long-tongued species. Funnel flowers (group 2) were visited mostly by long-tongued species, with medium- and short-tongued species following in rank order. The opposite pattern appears in head flowers (group 7), which were visited at the highest rate by short-tongued species, with medium- and long-tongued species following in reverse rank-order. Lip flowers (group 5) were visited at high rates by all tongue-length classes, with little differentiation between them. Flag flowers (group 6) were received similarly high rates of visitation from long- and medium-tongued species, with lower visitation from short-tongued species.

```{r fig3, echo = FALSE,  fig.cap = paste("Posterior conditional mean visits (log-scale) by Kugler morphotype and bumble bee tongue length class. Half-eye plots depict 66\\% and 95\\% CIs.")}
knitr::include_graphics("../output/brm_pvis11_plot1.png")
```

## 3.3. Consequences of bias for niche relationships
